---
title: "Beta_of_Cats"
author: "Frederik Hilbert"
date: "2026-05-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load-packages}
library(reshape2)
library(phyloseq)
library(vegan)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(grid)
```

```{r load-data}
wd <- getwd()
file_path <-  file.path(wd,"/data/molecular_ecology_17695.rds")

Cats <- readRDS(file_path)

Cats_ps <- Cats$phyloseq
Cats_ps
```

# Sanity check

```{r create-relative-abundance-object}
ps_rel <- transform_sample_counts(
  Cats_ps,
  function(x) x / sum(x)
)

sample_sums(ps_rel)
```

# Presence/absence object

```{r create-presence-absence-object}
ps_pa <- transform_sample_counts(
  Cats_ps,
  function(x) { ifelse(x > 0, 1, 0) }
)

sample_sums(ps_pa)
```

# Sample metadata

```{r extract-sample-metadata}
sample_df <- sample_data(ps_rel) %>%
  data.frame()

head(sample_df)
```

# 1. Bray-Curtis

## Bray-Curtis distance

```{r bray-distance}
bray_dist <- distance(
  ps_rel,
  method = "bray"
)

bray_dist
```

## Display Bray-Curtis PCoA

```{r pcoa-bray-basic, fig.width=8, fig.height=5}
ord_bray <- ordinate(
  ps_rel,
  method = "PCoA",
  distance = bray_dist
)

plot_ordination(
  ps_rel,
  ord_bray,
  shape = "origin",
  color = "location"
)
```

## Metadata values for Bray-Curtis PCoA

```{r bray-ordination-dataframe}
ord_bray_df <- plot_ordination(
  ps_rel,
  ord_bray,
  justDF = TRUE
)

head(ord_bray_df)
```

## Display Bray-Curtis with groups and ggplot

```{r pcoa-bray-with-groups, fig.width=8, fig.height=5}
ggplot(ord_bray_df, aes(x = Axis.1, y = Axis.2, color = location, shape = origin)) +
  geom_point(size = 3) +
  theme_bw()
```

## Eigenvalues for Bray-Curtis

```{r pcoa-bray-scree-plot, fig.width=8, fig.height=5}
plot_scree(ord_bray)
```

## Extracting Bray-Curtis ordination coordinates

```{r pcoa-bray-ggplot, fig.width=8, fig.height=5}
ggplot(ord_bray_df, aes(x = Axis.1, y = Axis.2, color = location)) +
  geom_point(size = 3) +
  theme_bw()
```

## NMDS based on Bray-Curtis distance

```{r nmds-bray-basic, fig.width=8, fig.height=5}
ord_nmds_bray <- ordinate(
  ps_rel,
  method = "NMDS",
  distance = bray_dist
)

plot_ordination(
  ps_rel,
  ord_nmds_bray,
  shape = "origin",
  color = "location"
)
```

## PERMANOVA Bray-Curtis by location

```{r permanova-bray-location}
adonis_bray_location <- adonis2(
  bray_dist ~ location,
  data = sample_df,
  permutations = 999
)

adonis_bray_location
```

## PERMANOVA Bray-Curtis by origin

```{r permanova-bray-origin}
adonis_bray_origin <- adonis2(
  bray_dist ~ origin,
  data = sample_df,
  permutations = 999
)

adonis_bray_origin
```

## Checking homogeneity of dispersion for Bray-Curtis by origin

```{r dispersion-bray-origin, fig.width=8, fig.height=5}
dispersion_bray_origin <- betadisper(
  bray_dist,
  sample_df$origin
)

anova(dispersion_bray_origin)
plot(dispersion_bray_origin)
```

## Checking homogeneity of dispersion for Bray-Curtis by location

```{r dispersion-bray-location, fig.width=8, fig.height=5}
dispersion_bray_location <- betadisper(
  bray_dist,
  sample_df$location
)

anova(dispersion_bray_location)
plot(dispersion_bray_location)
```

## PERMANOVA Bray-Curtis subset: Denmark vs Malaysia

```{r permanova-bray-subset}
ps_rel_bray_subset <- subset_samples(
  ps_rel,
  location %in% c("Spain", "Brazil")
)

bray_dist_subset <- distance(
  ps_rel_bray_subset,
  method = "bray"
)

sample_df_bray_subset <- sample_data(ps_rel_bray_subset) %>%
  data.frame()

adonis2(
  bray_dist_subset ~ location,
  data = sample_df_bray_subset,
  permutations = 999
)
```

# 2. Jaccard

## Jaccard distance

```{r jaccard-distance}
jaccard_dist <- distance(
  ps_pa,
  method = "jaccard"
)

jaccard_dist
```

## Display Jaccard PCoA

```{r pcoa-jaccard-basic, fig.width=8, fig.height=5}
ord_jaccard <- ordinate(
  ps_pa,
  method = "PCoA",
  distance = jaccard_dist
)

plot_ordination(
  ps_pa,
  ord_jaccard,
  shape = "origin",
  color = "location"
)
```

## Metadata values for Jaccard PCoA

```{r jaccard-ordination-dataframe}
ord_jaccard_df <- plot_ordination(
  ps_pa,
  ord_jaccard,
  justDF = TRUE
)

head(ord_jaccard_df)
```

## Display Jaccard with groups and ggplot

```{r pcoa-jaccard-with-groups, fig.width=8, fig.height=5}
ggplot(ord_jaccard_df, aes(x = Axis.1, y = Axis.2, color = location, shape = origin)) +
  geom_point(size = 3) +
  theme_bw()
```

## Eigenvalues for Jaccard

```{r pcoa-jaccard-scree-plot, fig.width=8, fig.height=5}
plot_scree(ord_jaccard)
```

## Extracting Jaccard ordination coordinates

```{r pcoa-jaccard-ggplot, fig.width=8, fig.height=5}
ggplot(ord_jaccard_df, aes(x = Axis.1, y = Axis.2, color = location)) +
  geom_point(size = 3) +
  theme_bw()
```

## NMDS based on Jaccard distance

```{r nmds-jaccard-basic, fig.width=8, fig.height=5}
ord_nmds_jaccard <- ordinate(
  ps_pa,
  method = "NMDS",
  distance = jaccard_dist
)

plot_ordination(
  ps_pa,
  ord_nmds_jaccard,
  shape = "origin",
  color = "location"
)
```

## PERMANOVA Jaccard by location

```{r permanova-jaccard-location}
adonis_jaccard_location <- adonis2(
  jaccard_dist ~ location,
  data = sample_df,
  permutations = 999
)

adonis_jaccard_location
```

## PERMANOVA Jaccard by origin

```{r permanova-jaccard-origin}
adonis_jaccard_origin <- adonis2(
  jaccard_dist ~ origin,
  data = sample_df,
  permutations = 999
)

adonis_jaccard_origin
```

## Checking homogeneity of dispersion for Jaccard by origin

```{r dispersion-jaccard-origin, fig.width=8, fig.height=5}
dispersion_jaccard_origin <- betadisper(
  jaccard_dist,
  sample_df$origin
)

anova(dispersion_jaccard_origin)
plot(dispersion_jaccard_origin)
```

## Checking homogeneity of dispersion for Jaccard by location

```{r dispersion-jaccard-location, fig.width=8, fig.height=5}
dispersion_jaccard_location <- betadisper(
  jaccard_dist,
  sample_df$location
)

anova(dispersion_jaccard_location)
plot(dispersion_jaccard_location)
```

## PERMANOVA Jaccard subset: Denmark vs Malaysia

```{r permanova-jaccard-subset}
ps_pa_jaccard_subset <- subset_samples(
  ps_pa,
  location %in% c("Denmark", "Malaysia")
)

jaccard_dist_subset <- distance(
  ps_pa_jaccard_subset,
  method = "jaccard"
)

sample_df_jaccard_subset <- sample_data(ps_pa_jaccard_subset) %>%
  data.frame()

adonis2(
  jaccard_dist_subset ~ location,
  data = sample_df_jaccard_subset,
  permutations = 999
)
```

# 3. Unweighted UniFrac

UniFrac requires a phylogenetic tree in the `phyloseq` object.

## Unweighted UniFrac distance

```{r unifrac-distance}
unifrac_dist <- distance(
  ps_pa,
  method = "unifrac",
  weighted = FALSE
)

unifrac_dist
```

## Display unweighted UniFrac PCoA

```{r pcoa-unifrac-basic, fig.width=8, fig.height=5}
ord_unifrac <- ordinate(
  ps_pa,
  method = "PCoA",
  distance = unifrac_dist
)

plot_ordination(
  ps_pa,
  ord_unifrac,
  shape = "origin",
  color = "location"
)
```

## Metadata values for unweighted UniFrac PCoA

```{r unifrac-ordination-dataframe}
ord_unifrac_df <- plot_ordination(
  ps_pa,
  ord_unifrac,
  justDF = TRUE
)

head(ord_unifrac_df)
```

## Display unweighted UniFrac with groups and ggplot

```{r pcoa-unifrac-with-groups, fig.width=8, fig.height=5}
ggplot(ord_unifrac_df, aes(x = Axis.1, y = Axis.2, color = location, shape = origin)) +
  geom_point(size = 3) +
  theme_bw()
```

## Eigenvalues for unweighted UniFrac

```{r pcoa-unifrac-scree-plot, fig.width=8, fig.height=5}
plot_scree(ord_unifrac)
```

## Extracting unweighted UniFrac ordination coordinates

```{r pcoa-unifrac-ggplot, fig.width=8, fig.height=5}
ggplot(ord_unifrac_df, aes(x = Axis.1, y = Axis.2, color = location)) +
  geom_point(size = 3) +
  theme_bw()
```

## NMDS based on unweighted UniFrac distance

```{r nmds-unifrac-basic, fig.width=8, fig.height=5}
ord_nmds_unifrac <- ordinate(
  ps_pa,
  method = "NMDS",
  distance = unifrac_dist
)

plot_ordination(
  ps_pa,
  ord_nmds_unifrac,
  shape = "origin",
  color = "location"
)
```

## PERMANOVA unweighted UniFrac by location

```{r permanova-unifrac-location}
adonis_unifrac_location <- adonis2(
  unifrac_dist ~ location,
  data = sample_df,
  permutations = 999
)

adonis_unifrac_location
```

## PERMANOVA unweighted UniFrac by origin

```{r permanova-unifrac-origin}
adonis_unifrac_origin <- adonis2(
  unifrac_dist ~ origin,
  data = sample_df,
  permutations = 999
)

adonis_unifrac_origin
```

## Checking homogeneity of dispersion for unweighted UniFrac by origin

```{r dispersion-unifrac-origin, fig.width=8, fig.height=5}
dispersion_unifrac_origin <- betadisper(
  unifrac_dist,
  sample_df$origin
)

anova(dispersion_unifrac_origin)
plot(dispersion_unifrac_origin)
```

## Checking homogeneity of dispersion for unweighted UniFrac by location

```{r dispersion-unifrac-location, fig.width=8, fig.height=5}
dispersion_unifrac_location <- betadisper(
  unifrac_dist,
  sample_df$location
)

anova(dispersion_unifrac_location)
plot(dispersion_unifrac_location)
```

## PERMANOVA unweighted UniFrac subset: Denmark vs Malaysia

```{r permanova-unifrac-subset}
ps_pa_unifrac_subset <- subset_samples(
  ps_pa,
  location %in% c("Brazil", "Spain")
)

unifrac_dist_subset <- distance(
  ps_pa_unifrac_subset,
  method = "unifrac",
  weighted = FALSE
)

sample_df_unifrac_subset <- sample_data(ps_pa_unifrac_subset) %>%
  data.frame()

adonis2(
  unifrac_dist_subset ~ location,
  data = sample_df_unifrac_subset,
  permutations = 999
)
```

# 4. Weighted UniFrac

UniFrac requires a phylogenetic tree in the `phyloseq` object.

## Weighted UniFrac distance

```{r weighted-unifrac-distance}
wunifrac_dist <- distance(
  ps_rel,
  method = "unifrac",
  weighted = TRUE
)

wunifrac_dist
```

## Display weighted UniFrac PCoA

```{r pcoa-weighted-unifrac-basic, fig.width=8, fig.height=5}
ord_wunifrac <- ordinate(
  ps_rel,
  method = "PCoA",
  distance = wunifrac_dist
)

plot_ordination(
  ps_rel,
  ord_wunifrac,
  shape = "origin",
  color = "location"
)
```

## Metadata values for weighted UniFrac PCoA

```{r weighted-unifrac-ordination-dataframe}
ord_wunifrac_df <- plot_ordination(
  ps_rel,
  ord_wunifrac,
  justDF = TRUE
)

head(ord_wunifrac_df)
```

## Display weighted UniFrac with groups and ggplot

```{r pcoa-weighted-unifrac-with-groups, fig.width=8, fig.height=5}
ggplot(ord_wunifrac_df, aes(x = Axis.1, y = Axis.2, color = location, shape = origin)) +
  geom_point(size = 3) +
  theme_bw()
```

## Eigenvalues for weighted UniFrac

```{r pcoa-weighted-unifrac-scree-plot, fig.width=8, fig.height=5}
plot_scree(ord_wunifrac)
```

## Extracting weighted UniFrac ordination coordinates

```{r pcoa-weighted-unifrac-ggplot, fig.width=8, fig.height=5}
ggplot(ord_wunifrac_df, aes(x = Axis.1, y = Axis.2, color = location)) +
  geom_point(size = 3) +
  theme_bw()
```

## NMDS based on weighted UniFrac distance

```{r nmds-weighted-unifrac-basic, fig.width=8, fig.height=5}
ord_nmds_wunifrac <- ordinate(
  ps_rel,
  method = "NMDS",
  distance = wunifrac_dist
)

plot_ordination(
  ps_rel,
  ord_nmds_wunifrac,
  shape = "origin",
  color = "location"
)
```

## PERMANOVA weighted UniFrac by location

```{r permanova-weighted-unifrac-location}
adonis_wunifrac_location <- adonis2(
  wunifrac_dist ~ origin,
  data = sample_df,
  permutations = 999
)

adonis_wunifrac_location
```

## PERMANOVA weighted UniFrac by origin

```{r permanova-weighted-unifrac-origin}
adonis_wunifrac_origin <- adonis2(
  wunifrac_dist ~ origin,
  data = sample_df,
  permutations = 999
)

adonis_wunifrac_origin
```

## Checking homogeneity of dispersion for weighted UniFrac by origin

```{r dispersion-weighted-unifrac-origin, fig.width=8, fig.height=5}
dispersion_wunifrac_origin <- betadisper(
  wunifrac_dist,
  sample_df$origin
)

anova(dispersion_wunifrac_origin)
plot(dispersion_wunifrac_origin)
```

## Checking homogeneity of dispersion for weighted UniFrac by location

```{r dispersion-weighted-unifrac-location, fig.width=8, fig.height=5}
dispersion_wunifrac_location <- betadisper(
  wunifrac_dist,
  sample_df$location
)

anova(dispersion_wunifrac_location)
plot(dispersion_wunifrac_location)
```

## PERMANOVA weighted UniFrac subset: Denmark vs Malaysia

```{r permanova-weighted-unifrac-subset}
ps_rel_wunifrac_subset <- subset_samples(
  ps_rel,
  location %in% c("Denmark", "Malaysia")
)

wunifrac_dist_subset <- distance(
  ps_rel_wunifrac_subset,
  method = "unifrac",
  weighted = TRUE
)

sample_df_wunifrac_subset <- sample_data(ps_rel_wunifrac_subset) %>%
  data.frame()

adonis2(
  wunifrac_dist_subset ~ location,
  data = sample_df_wunifrac_subset,
  permutations = 999
)
```

# Compare all beta-diversity plots together

The following sections place the plots from each beta-diversity method into 2x2 layouts so the four methods can be compared directly.

```{r helper-function-for-2x2-ggplots}
# Helper function for arranging four ggplot objects in a 2x2 layout
plot_2x2 <- function(p1, p2, p3, p4) {
  grid.newpage()
  pushViewport(viewport(layout = grid.layout(2, 2)))
  print(p1, vp = viewport(layout.pos.row = 1, layout.pos.col = 1))
  print(p2, vp = viewport(layout.pos.row = 1, layout.pos.col = 2))
  print(p3, vp = viewport(layout.pos.row = 2, layout.pos.col = 1))
  print(p4, vp = viewport(layout.pos.row = 2, layout.pos.col = 2))
}
```

## PCoA plots with location and origin

```{r compare-all-methods-2x2-pcoa-location-origin, fig.width=12, fig.height=10}
bray_pcoa_plot <- ggplot(ord_bray_df, aes(x = Axis.1, y = Axis.2, color = location, shape = origin)) +
  geom_point(size = 3) +
  theme_bw() +
  labs(title = "Bray-Curtis", x = "PCoA Axis 1", y = "PCoA Axis 2")

jaccard_pcoa_plot <- ggplot(ord_jaccard_df, aes(x = Axis.1, y = Axis.2, color = location, shape = origin)) +
  geom_point(size = 3) +
  theme_bw() +
  labs(title = "Jaccard", x = "PCoA Axis 1", y = "PCoA Axis 2")

unifrac_pcoa_plot <- ggplot(ord_unifrac_df, aes(x = Axis.1, y = Axis.2, color = location, shape = origin)) +
  geom_point(size = 3) +
  theme_bw() +
  labs(title = "Unweighted UniFrac", x = "PCoA Axis 1", y = "PCoA Axis 2")

weighted_unifrac_pcoa_plot <- ggplot(ord_wunifrac_df, aes(x = Axis.1, y = Axis.2, color = location, shape = origin)) +
  geom_point(size = 3) +
  theme_bw() +
  labs(title = "Weighted UniFrac", x = "PCoA Axis 1", y = "PCoA Axis 2")

plot_2x2(
  bray_pcoa_plot,
  jaccard_pcoa_plot,
  unifrac_pcoa_plot,
  weighted_unifrac_pcoa_plot
)
```

## PCoA plots with location only

```{r compare-all-methods-2x2-pcoa-location-only, fig.width=12, fig.height=10}
bray_pcoa_location_plot <- ggplot(ord_bray_df, aes(x = Axis.1, y = Axis.2, color = location)) +
  geom_point(size = 3) +
  theme_bw() +
  labs(title = "Bray-Curtis", x = "PCoA Axis 1", y = "PCoA Axis 2")

jaccard_pcoa_location_plot <- ggplot(ord_jaccard_df, aes(x = Axis.1, y = Axis.2, color = location)) +
  geom_point(size = 3) +
  theme_bw() +
  labs(title = "Jaccard", x = "PCoA Axis 1", y = "PCoA Axis 2")

unifrac_pcoa_location_plot <- ggplot(ord_unifrac_df, aes(x = Axis.1, y = Axis.2, color = location)) +
  geom_point(size = 3) +
  theme_bw() +
  labs(title = "Unweighted UniFrac", x = "PCoA Axis 1", y = "PCoA Axis 2")

weighted_unifrac_pcoa_location_plot <- ggplot(ord_wunifrac_df, aes(x = Axis.1, y = Axis.2, color = location)) +
  geom_point(size = 3) +
  theme_bw() +
  labs(title = "Weighted UniFrac", x = "PCoA Axis 1", y = "PCoA Axis 2")

plot_2x2(
  bray_pcoa_location_plot,
  jaccard_pcoa_location_plot,
  unifrac_pcoa_location_plot,
  weighted_unifrac_pcoa_location_plot
)
```

## Eigenvalue scree plots

```{r compare-all-methods-2x2-scree, fig.width=12, fig.height=10}
bray_scree_plot <- plot_scree(ord_bray) +
  ggtitle("Bray-Curtis")

jaccard_scree_plot <- plot_scree(ord_jaccard) +
  ggtitle("Jaccard")

unifrac_scree_plot <- plot_scree(ord_unifrac) +
  ggtitle("Unweighted UniFrac")

weighted_unifrac_scree_plot <- plot_scree(ord_wunifrac) +
  ggtitle("Weighted UniFrac")

plot_2x2(
  bray_scree_plot,
  jaccard_scree_plot,
  unifrac_scree_plot,
  weighted_unifrac_scree_plot
)
```

## NMDS plots

```{r compare-all-methods-2x2-nmds, fig.width=12, fig.height=10}
bray_nmds_plot <- plot_ordination(
  ps_rel,
  ord_nmds_bray,
  shape = "origin",
  color = "location"
) +
  geom_point(size = 3) +
  theme_bw() +
  labs(title = "Bray-Curtis")

jaccard_nmds_plot <- plot_ordination(
  ps_pa,
  ord_nmds_jaccard,
  shape = "origin",
  color = "location"
) +
  geom_point(size = 3) +
  theme_bw() +
  labs(title = "Jaccard")

unifrac_nmds_plot <- plot_ordination(
  ps_pa,
  ord_nmds_unifrac,
  shape = "origin",
  color = "location"
) +
  geom_point(size = 3) +
  theme_bw() +
  labs(title = "Unweighted UniFrac")

weighted_unifrac_nmds_plot <- plot_ordination(
  ps_rel,
  ord_nmds_wunifrac,
  shape = "origin",
  color = "location"
) +
  geom_point(size = 3) +
  theme_bw() +
  labs(title = "Weighted UniFrac")

plot_2x2(
  bray_nmds_plot,
  jaccard_nmds_plot,
  unifrac_nmds_plot,
  weighted_unifrac_nmds_plot
)
```

## Dispersion plots by origin

```{r compare-all-methods-2x2-dispersion-origin, fig.width=12, fig.height=10}
old_par <- par(mfrow = c(2, 2))

plot(dispersion_bray_origin, main = "Bray-Curtis: origin")
plot(dispersion_jaccard_origin, main = "Jaccard: origin")
plot(dispersion_unifrac_origin, main = "Unweighted UniFrac: origin")
plot(dispersion_wunifrac_origin, main = "Weighted UniFrac: origin")

par(old_par)
```

## Dispersion plots by location

```{r compare-all-methods-2x2-dispersion-location, fig.width=12, fig.height=10}
old_par <- par(mfrow = c(2, 2))

plot(dispersion_bray_location, main = "Bray-Curtis: location")
plot(dispersion_jaccard_location, main = "Jaccard: location")
plot(dispersion_unifrac_location, main = "Unweighted UniFrac: location")
plot(dispersion_wunifrac_location, main = "Weighted UniFrac: location")

par(old_par)
```

# Create pairwise permanova matrix small scripts
```{r}
pairwise_permanova_matrix <- function(dist_obj, metadata, group_var, groups_keep) {
  
  # Ensure rownames exist
  metadata <- metadata
  metadata <- metadata[rownames(metadata) %in% rownames(as.matrix(dist_obj)), ]
  
  # Subset metadata correctly
  meta_sub <- metadata %>%
    dplyr::filter(.data[[group_var]] %in% groups_keep) %>%
    droplevels()
  
  # Subset distance matrix using matching samples
  dist_mat <- as.matrix(dist_obj)
  dist_sub <- dist_mat[rownames(meta_sub), rownames(meta_sub)]
  
  groups <- unique(meta_sub[[group_var]])
  
  results <- data.frame()
  
  for (i in 1:(length(groups)-1)) {
    for (j in (i+1):length(groups)) {
      
      g1 <- groups[i]
      g2 <- groups[j]
      
      pair_meta <- meta_sub %>%
        dplyr::filter(.data[[group_var]] %in% c(g1, g2))
      
      pair_samples <- rownames(pair_meta)
      pair_dist <- dist_sub[pair_samples, pair_samples]
      pair_dist <- as.dist(pair_dist)
      
      # Build formula safely
      form <- as.formula(paste("pair_dist ~", group_var))
      
      perm <- adonis2(
        form,
        data = pair_meta,
        permutations = 999
      )
      
      results <- rbind(
        results,
        data.frame(
          group1 = g1,
          group2 = g2,
          R2 = perm$R2[1],
          p = perm$`Pr(>F)`[1]
        )
      )
    }
  }
  
  # Adjust p-values
  results$p_adj <- p.adjust(results$p, method = "BH")
  
  # Create symmetric matrix
  mat <- matrix(NA, nrow = length(groups), ncol = length(groups))
  rownames(mat) <- colnames(mat) <- groups
  
  for (k in 1:nrow(results)) {
    mat[results$group1[k], results$group2[k]] <- results$p_adj[k]
    mat[results$group2[k], results$group1[k]] <- results$p_adj[k]
  }
  
  return(list(table = results, matrix = mat))
}

```
## Locations in vector
```{r}
locations_of_interest <- c(
  "Spain", "Brazil", "Malaysia",
  "Denmark", "Carbo Verde", "Aruba"
)
```
## Pairwise test for bray dist
```{r}
pairwise_bray <- pairwise_permanova_matrix(
  bray_dist,
  sample_df,
  "location",
  locations_of_interest
)

pairwise_bray$table
pairwise_bray$matrix
```
## Pairwise test for Jaccard
```{r}
pairwise_jaccard <- pairwise_permanova_matrix(
  jaccard_dist,
  sample_df,
  "location",
  locations_of_interest
)

pairwise_jaccard$table
pairwise_jaccard$matrix
```
## Pairwise test for unweighted UniFrac
```{r}
pairwise_unifrac <- pairwise_permanova_matrix(
  unifrac_dist,
  sample_df,
  "location",
  locations_of_interest
)

pairwise_unifrac$table
pairwise_unifrac$matrix
```
## Pairwise test for weighted UniFrac
```{r}
pairwise_wunifrac <- pairwise_permanova_matrix(
  wunifrac_dist,
  sample_df,
  "location",
  locations_of_interest
)

pairwise_wunifrac$table
pairwise_wunifrac$matrix
```
## Print all matrix in one
```{r}
plot_matrix <- function(mat, title) {
  df <- melt(mat, na.rm = TRUE)
  ggplot(df, aes(Var1, Var2, fill = value)) +
    geom_tile() +
    geom_text(aes(label = round(value, 3))) +
    scale_fill_gradient(low = "white", high = "red") +
    theme_minimal() +
    labs(title = title, x = "", y = "", fill = "Adj p")
}

p1 <- plot_matrix(pairwise_bray$matrix, "Bray-Curtis PERMANOVA")
p2 <- plot_matrix(pairwise_jaccard$matrix, "Jaccard PERMANOVA")
p3 <- plot_matrix(pairwise_unifrac$matrix, "Unweighted UniFrac PERMANOVA")
p4 <- plot_matrix(pairwise_wunifrac$matrix, "Weighted UniFrac PERMANOVA")

grid.arrange(p1, p2, p3, p4, ncol = 2)

```


