---
title: "Meta project about cats"
author: "Ellen Bendtsen"
email: "cxk650@ku.dk"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Loading of required libraries

```{r}
library(phyloseq)
library(vegan)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(ggplot2)
library(tidyverse)
library(ggh4x)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(microbiome)
library(glue)
```

## Load and inspect data

```{r}
obj <- readRDS("C:/Users/emtb3/Downloads/molecular_ecology_17695.rds")
```

# Inspect the object:
```{r}
class(obj)
names(obj)
str(obj, max.level = 2)
```

The .rds file stores several elements. The phyloseq object is extracted from the list:
```{r}
cat_data <- obj$phyloseq
```

The phyloseq object contains an OTU table, sample metadata, a taxonomy table and a phylogenetic tree. Let's inspect these and the data:
```{r}
sample_variables(cat_data) # origin & location
rank_names(cat_data)       # ranking for the taxa
slotNames(cat_data)        # elements stored in the phyloseq object

cat_data                   # overview of data
```
We can see that the two sample variables in the dataset are:
- 'origin': whether cats are domestic or feral
- 'location': country/location of sampling.

Furthermore, we have found 214 different taxa across 92 samples.

# Further exploration of the phyloseq object:
```{r}
# This stores the metadata about the samples themselves. E.g. where the samples came from (location) and domestic vs. feral (origin).

head(sample_data(cat_data))

# This stores the taxonomy/classification of each taxon/MAG. It links each row in the OTU table to biological names.
# So when OTU says "bin_1 has abundance 45" tax table says "bin_1 belongs to Lactobacillus". Typically used for grouping taxa at genus level, coloring taxa in plots, filtering taxa, biological interpretation...

head(tax_table(cat_data))

# The main quantitative data matrix. This stores how much of each microbe/MAG/taxon is present in each sample. Rows are usually taxa/MAGS and columns are usually samples. Use this for diversity analysis, PCA, differential abundance...

head(otu_table(cat_data))
```
# Convert phyloseq components to dataframes

These conversions make it easier to inspect and manipulate the data with tidyverse functions.

Convert otu_table to a data frame:
```{r}
otu <- otu_table(cat_data, errorIfNULL = FALSE)

if (!is.null(otu)) {
  otu_df <- as.data.frame(as(otu, "matrix"))
  View(otu_df)
}
```

Convert sample_data to data frame:
```{r}
sam <- sample_data(cat_data, errorIfNULL = FALSE)

if (!is.null(sam)) {
  sam_df <- as.data.frame(sam)
  View(sam_df)
}
```

Convert taxonomy_table to data frame:
```{r}
tax <- tax_table(cat_data, errorIfNULL = FALSE)

if (!is.null(tax)) {
  tax_df <- as.data.frame(as(tax, "matrix"))
  View(tax_df)
}
```


# Inspect and visualize the phylogenetic tree:

Inspect the phy_tree object
```{r}
tree <- phy_tree(cat_data)
tree
```
Visualize the phylogenetic tree:
```{r, fig.width=10, fig.height=10}
plot(tree, type = "fan", cex = 0.35)
```
For a tree with this many tips, it is often easier to save it as a large PDF and zoom in:
```{r, eval=FALSE}
pdf("cat_phylogenetic_tree.pdf", width = 20, height = 20)
plot(tree, type = "fan", cex = 0.35)
dev.off()
```

## Community composition

# Relative abundance transformation

The raw abundance table is transformed to relative abundance so that samples can be compared despite differences in sequencing depth. 

This is necessary because microbiome samples may contain different total numbers of reads. If we compared raw counts directly, a sample with more sequencing reads could look like it had more bacteria simply because it was sequenced more deeply.

Relative abundance therefore asks: out of the total microbiome profile in each sample, what proportion belongs to each taxon? This makes samples more comparable for exploratory plots such as stacked barplots.

However, relative abundance data are compositional, meaning that all taxa within one sample sum to 1. If one taxon increases, others must decrease proportionally, even if their absolute abundance did not biologically change. This is why later analyses also include CLR-transformed data.

# Genus-level composition

The genus-level stacked barplot is used as an exploratory overview of microbiome composition. We aggregate taxa to genus level, because plotting all individual bins/MAGs would be too detailed and difficult to interpret visually. Genus-level aggregation gives a broader biological summary while still retaining more detail than phylum-level plots.

The purpose of this plot is not formal hypothesis testing. Instead, it helps us visually inspect whether domestic and feral cats, or cats from different locations, appear to have different dominant microbial groups.

```{r}
cat_rel <- transform_sample_counts(cat_data, function(x) x / sum(x))

otu_rel <- as(otu_table(cat_rel), "matrix") %>% t() %>% as.data.frame()

meta_rel <- as(sample_data(cat_rel), "matrix") %>% as.data.frame()

tax_rel <- as(tax_table(cat_rel), "matrix") %>% as.data.frame()

cat_genus <- tax_glom(cat_rel, taxrank = "genus")

plot_bar(cat_genus, fill = "genus") +
  facet_nested(~ location + origin, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_blank())
```

# Sample overview

Before statistical testing, it is useful to check whether the sampling design is balanced. This is especially important here, because both 'origin' and 'location' may influence the gut microbiome.

The main biological comparison in this project is domestic versus feral cats. However, the samples also come from different locations. If domestic and feral cats are not evenly represented across locations, then the origin and geography may be partly confounded. For example, if many feral cats come from one country and many domestic cats from another, an apparant domestic/feral difference could actually reflect geography, diet environment or local sampling effects.

Therefore, the following plot is used to check how many domestic and feral cats are available from each location before interpreting the statistical tests.
```{r}
meta_rel %>%
  count(location, origin) %>%
  ggplot(aes(x = location, y = n, fill = origin)) +
  geom_col(position = "dodge") +
  labs(
    x = "Location",
    y = "Number of samples",
    title = "Sampling design: number of domestic and feral cats per location"
  ) +
  theme_bw()

meta_rel %>%
  count(location, origin)
```

## Create long data frame for downstream analyses.

The chunk below creates `dat_rel`, the long relative-abundance table used by the remaining relative-abundance analyses
```{r}
dat_rel <- otu_rel %>%
  rownames_to_column("sample_id") %>%
  pivot_longer(-sample_id, names_to = "taxon", values_to = "rel_abund") %>%
  left_join(meta_rel %>% as("data.frame") %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  left_join(tax_rel %>% rownames_to_column("taxon"), by = "taxon")

head(dat_rel)
```

## Diagnose the data before testing

Microbiome relative abundance data often contains many zeros and strongly right-skewed distributions. These diagnostic plots help justify using non-parametric tests and log/CLR transformed models.

This step is included because the choice of statistical test should depend on the structure of the data. Many standard parametric tests assume approximately normally distributed residuals and relatively even variance. However, microbiome data often violate these assumptions because many taxa are absent from mane samples, whilea few taxa can be highly abundant in only a small number of samples. (måske find et eksempel på vores taxa fordelings plot)

The diagnostic plots are therefore not meant to answer the biological question directly. They are used to explain why robust and compositional methods are more appropriate than simple normal-theory tests on raw relative abundance.

# Zero inflation:
```{r}
dat_rel %>%
  group_by(taxon) %>%
  summarise(prop_zero = mean(rel_abund == 0), .groups = "drop") %>%
  ggplot(aes(x = prop_zero)) +
  geom_histogram(binwidth = 0.05, fill = "steelblue", color = "white") +
  labs(
    x = "Proportion of samples where relative abundance = 0",
    y = "Number of taxa",
    title = "Zero-inflation across taxa"
  ) +
  theme_bw()
```
High zero proportions mean that many taxa are absent from many samples, which limits information for per-taxon testing.

Our plot shows us that most taxa are absent from most samples.

In the original script, we could filter according to origin or location. Why and why not do this. We don't do it now right? I think we just don't filter and look at all taxa at once and not just the ones in domestic or feral cats or in a specific location.

# Skewness in selected taxa
```{r}
set.seed(42)

example_taxa <- dat_rel %>%
  group_by(taxon, genus, species) %>%
  summarise(prop_nonzero = mean(rel_abund > 0), .groups = "drop") %>%
  filter(prop_nonzero > 0.4, prop_nonzero < 0.9) %>%
  slice_sample(n = 6) %>%
  pull(taxon)

dat_rel %>%
  filter(taxon %in% example_taxa, rel_abund > 0) %>%
  mutate(label = ifelse(is.na(species), taxon, species)) %>%
  ggplot(aes(x = rel_abund)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  facet_wrap(~ label, scales = "free") +
  labs(
    x = "Relative abundance (non-zero values only)",
    y = "Count",
    title = "Right-skewed distributions in selected cat gut taxa"
  ) +
  theme_bw()
```
What difference does this part make: "ifelse(is.na(species), taxon, species)" vs just "species"?

What do these plots tell us?
- "Use these histograms to see whether non-zero abundances are dominated by a few high values."?

# Q-Q plots
```{r}
dat_rel %>%
  filter(taxon %in% example_taxa) %>%
  mutate(label = ifelse(is.na(species), taxon, species)) %>%
  ggplot(aes(sample = rel_abund)) +
  stat_qq(size = 0.8, alpha = 0.6) +
  stat_qq_line(color = "red", linetype = "dashed") +
  facet_wrap(~ label, scales = "free") +
  labs(
    x = "Theoretical quantiles (normal)",
    y = "Sample quantiles",
    title = "Q-Q plots: departures from normality in selected taxa"
  ) +
  theme_bw()
```
Use the Q-Q plots as a visual check of normality on raw relative abundances?

These plots are not final results, but they support the choice of robust/non.parametric and compositional methods.

## Explore individual taxa

Both by 'origin' and by 'location'

This section analyses one taxon at a time. Because many taxa are tested, p-values are adjusted using the Benjamini-Hochberg false discovery rate method. - what is this? I thought we had good reason not to look at individual taxon?

Change `one_taxon` to inspect another taxon before interpreting formal results.
- why 1e-6?
- why alpha = 0.6?
- argument for log skala?
```{r}
one_taxon <- dat_rel$taxon[1]

# Comparison between origins
dat_rel %>%
  filter(taxon == one_taxon) %>%
  ggplot(aes(x = origin, y = rel_abund + 1e-6, fill = origin)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.6) +
    scale_y_log10() +
    labs(
      x = 'Origin',
      y = 'Realtive abundance + 1e-6 (log10 scale)',
      title = glue('Relative abundance of {one_taxon} by origin')
    )
    theme_bw()
```
```{r}
dat_rel %>%
  filter(taxon == one_taxon) %>%
  ggplot(aes(x = location, y = rel_abund + 1e-6, fill = location)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6) +
  scale_y_log10() +
  labs(
    x = "Location",
    y = "Relative abundance + 1e-6 (log10 scale)",
    title = glue("Relative abundance of {one_taxon} by location")
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Wilcoxon rank-sum test: domestic vs. feral
- why na.rm = TRUE?
- what is log2 fold change?
- why not log10?
- what excatly does the Wilcoxon rank-sum test do? Maybe I should look at the slides.

Our Wilcoxon rank-sum test compares relative abundance between two origins: domestic and feral. This is a simple, non-parametric test that does not assume normally distributed data, which makes it useful for zero-inflated and skewed microbiome abundances.

This test is included as a first-pass taxon-level comparison. It asks whether each taxon tends to have higher or lower relative abundance in domestic cats compared with feral cats.

A limitation is that this test does not adjust for location. Therefore, a significant result could reflect a true domestic/feral difference, but it could also partly reflect geographical differences if origin and location are not balanced.

```{r}
wilcox_origin_results <- dat_rel %>%
  group_by(taxon, genus, species) %>%
  summarise(
    p_value = wilcox.test(rel_abund ~ origin, exact = FALSE)$p.value,
    mean_domestic = mean(rel_abund[origin == "Domestic"], na.rm = TRUE),
    mean_feral = mean(rel_abund[origin == "Feral"], na.rm = TRUE),
    log2_fc_feral_vs_domestic = log2((mean_feral + 1e-6) / (mean_domestic + 1e-6)),
    .groups = "drop"
  ) %>%
  mutate(q_value = p.adjust(p_value, method = "BH"))

wilcox_origin_results %>%
  arrange(q_value) %>%
  head(20)
```

Positive log2 fold change means the taxon is enriched in feral cats.
Negative log2 fold change means it is enriched in domestic cats

```{r}
wilcox_origin_results %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & log2_fc_feral_vs_domestic > 0 ~ "Feral-enriched",
      q_value < 0.05 & log2_fc_feral_vs_domestic < 0 ~ "Domestic-enriched",
      TRUE ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = log2_fc_feral_vs_domestic, y = neg_log10_q, color = direction)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(
    x = "Log2 fold change (feral vs domestic)",
    y = "-log10(q-value)",
    color = NULL,
    title = "Wilcoxon rank-sum test: domestic vs feral"
  ) +
  theme_bw()
```
- what do you mean 'Not significant'?
- is this a volcano plot?
- is each dot a taxon, adn if it is enriched in domestic cats it is negativ and in feral cats it is positive?

The next plot shows Wilcoxon effect size on the x-axis and orders taxa by q-value.

```{r}
# Bar chart of top significant taxa ordered by q-value (Wilcoxon)
bind_rows(
  wilcox_origin_results %>% filter(q_value < 0.05) %>% arrange(log2_fc) %>% head(10),
  wilcox_origin_results %>% filter(q_value < 0.05) %>% arrange(-log2_fc) %>% head(10)
) %>%
  distinct() %>%
  mutate(
    label = species,
    direction = ifelse(log2_fc > 0, "Domestic-enriched", "Feral-enriched")
  ) %>%
  ggplot(aes(x = reorder(label, -q_value), y = log2_fc, fill = direction)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("Domestic-enriched" = "#E74C3C", "Feral-enriched" = "#3498DB")) +
    labs(x = NULL, y = "Log2 fold change (domestic vs feral)", fill = NULL,
         title = "Top differentially abundant taxa (Wilcoxon)") +
    theme_bw()
```

## Kruskal-Wallis test: differences between locations

The kruskal-Wallis test compares relative abundance across more than two groups. Here it is used to test whether taxa differ between locations.

This is relevant because the cats were sampled from several countries. Location may influence the gut microbiome through diet, enivronment, climate, host population structure or local sampling differences. Testing location seperately helps us evaluate whether geography is an important source of microbiome variation.

This test is the multi-group equivalent of a non-parametric comparison. It tells us whether there is evidence that a taxon differs across locations, but it does not directly identify which specific pair of locations differ.

```{r}
kruskal_location_results <- dat_rel %>%
  group_by(taxon, genus, species) %>%
  filter(n_distinct(location) > 1) %>%
  summarise(
    p_value = kruskal.test(rel_abund ~ location)$p.value,
    .groups = "drop"
  ) %>%
  mutate(q_value = p.adjust(p_value, method = "BH"))

kruskal_location_results %>%
  arrange(q_value) %>%
  head(20)
```
This test says whether location matters for a taxon, but it does not say which specific locations differ from each other.

## Log-transformed linear models

The next models test the domestic/feral effect while adjusting for location. This is important because microbiome differences may reflect geography rather than  origin alone.

Compared with the Wilcoxon test, the linear model is more flexible because it can include more than one explanatory variable. Here 'origin' is the main variable of interest, while 'location' is included as an adjustment variable.

The response variable is 'log10(rel_abund + 1e-6)'. The small pseudocount allows zeros to be log-transformed. The log transformation reduces the influence of extreme high values and makes skewed abundance distributions easier to model.

# GLM: origin adjusted for location
- why   filter(n_distinct(origin) == 2, n_distinct(location) > 1) %>%

With location as main variable and every single country as reference country:

```{r}
reference_locations <- sort(unique(dat_rel$location))

glm_all_references_results <- map_dfr(reference_locations, function(ref_location) {
  
  dat_rel %>%
    mutate(
      origin = relevel(factor(origin), ref = "Domestic"),
      location = relevel(factor(location), ref = ref_location)
    ) %>%
    group_by(taxon, genus, species) %>%
    filter(n_distinct(origin) == 2, n_distinct(location) > 1) %>%
    group_modify(~ {
      fit <- glm(
        log10(rel_abund + 1e-6) ~ location + origin,
        data = .x,
        family = gaussian()
      )
      
      broom::tidy(fit)
    }) %>%
    ungroup() %>%
    filter(str_detect(term, "^location")) %>%
    mutate(
      reference_location = ref_location
    )
}) %>%
  group_by(reference_location) %>%
  mutate(
    q_value = p.adjust(p.value, method = "BH")
  ) %>%
  ungroup()
```

```{r}
glm_all_references_results %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & estimate > 0 ~ "Higher than reference location",
      q_value < 0.05 & estimate < 0 ~ "Lower than reference location",
      TRUE ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = estimate, y = neg_log10_q, color = direction)) +
  geom_point(alpha = 0.45, size = 1.4) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  facet_wrap(~ reference_location, scales = "free_y") +
  labs(
    x = "GLM estimate (log10 relative abundance, location vs reference)",
    y = "-log10(q-value)",
    color = NULL,
    title = "Location effects using each country as reference",
    subtitle = "Each panel compares all other locations to the reference location, adjusted for origin"
  ) +
  theme_bw()
```

## Compositional analysis with CLR transformation with location as main variable and adjustment for origin

Microbiome data are compositional because only the relative proportions are observed. CLR transormation is used as a compositionally aware approach.

This matters because an increase in one taxon automatically forces the relative abundance of other taxon downward, even if their absolute abundances did not actually decrease. CLR transformation helps analyse taxa relative to the overall microbial community composition rather than treating each relative abundance as fully independent.

The CLR model is therefore included as a more compositionally appropriate version of the taxon-level analysis. In this project, it tests whether taxa are associated with domestic or feral origin after accounting for location.
```{r}
locations <- sort(unique(dat_clr$location))

clr_location_vs_rest_results <- map_dfr(locations, function(this_location) {
  
  dat_clr %>%
    mutate(
      location_group = ifelse(location == this_location, this_location, "Other"),
      location_group = factor(location_group, levels = c("Other", this_location)),
      origin = relevel(factor(origin), ref = "Domestic")
    ) %>%
    group_by(taxon, genus, species) %>%
    group_modify(~ {
      fit <- lm(
        clr ~ location_group + origin,
        data = .x
      )
      
      broom::tidy(fit)
    }) %>%
    ungroup() %>%
    filter(str_detect(term, "^location_group")) %>%
    mutate(
      target_location = this_location
    )
}) %>%
  group_by(target_location) %>%
  mutate(q_value = p.adjust(p.value, method = "BH")) %>%
  ungroup()
```

```{r}
clr_location_vs_rest_results %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & estimate > 0 ~ "Enriched in target location",
      q_value < 0.05 & estimate < 0 ~ "Depleted in target location",
      TRUE ~ "Not significant"
    ),
    label = ifelse(is.na(species), taxon, species)
  ) %>%
  ggplot(aes(x = estimate, y = neg_log10_q, color = direction)) +
  geom_point(alpha = 0.6, size = 1.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  facet_wrap(~ target_location, scales = "free_y") +
  labs(
    x = "CLR estimate (target location vs all other locations)",
    y = "-log10(q-value)",
    color = NULL,
    title = "Location-specific CLR volcano plots",
    subtitle = "Each panel compares one country to all others, adjusted for origin"
  ) +
  theme_bw()
```
## RDS:
```{r}
locations <- sort(unique(dat_rel$location))

rel_location_vs_rest_results <- map_dfr(locations, function(this_location) {
  
  dat_rel %>%
    mutate(
      location_group = ifelse(location == this_location, this_location, "Other"),
      location_group = factor(location_group, levels = c("Other", this_location)),
      origin = relevel(factor(origin), ref = "Domestic")
    ) %>%
    group_by(taxon, genus, species) %>%
    group_modify(~ {
      fit <- lm(
        log10(rel_abund + 1e-6) ~ location_group + origin,
        data = .x
      )
      
      broom::tidy(fit)
    }) %>%
    ungroup() %>%
    filter(str_detect(term, "^location_group")) %>%
    mutate(
      target_location = this_location
    )
}) %>%
  group_by(target_location) %>%
  mutate(q_value = p.adjust(p.value, method = "BH")) %>%
  ungroup()
```

```{r}
rel_location_vs_rest_results %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & estimate > 0 ~ "Enriched in target location",
      q_value < 0.05 & estimate < 0 ~ "Depleted in target location",
      TRUE ~ "Not significant"
    ),
    label = ifelse(is.na(species), taxon, species)
  ) %>%
  ggplot(aes(x = estimate, y = neg_log10_q, color = direction)) +
  geom_point(alpha = 0.6, size = 1.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  facet_wrap(~ target_location, scales = "free_y") +
  labs(
    x = "Estimate (log10 relative abundance, target location vs all others)",
    y = "-log10(q-value)",
    color = NULL,
    title = "Location-specific volcano plots based on log-relative abundance",
    subtitle = "Each panel compares one country to all others, adjusted for origin"
  ) +
  theme_bw()
```

