---
title: "Reproduced analyses from project report"
output: html_document
date: "2026-06-04"
---

```{r setup, include=FALSE}
# ---------------------------
# AUTO-INSTALL MISSING PACKAGES
# ---------------------------
required_packages <- c(
  "phyloseq", "vegan", "dplyr", "tibble", "purrr", "stringr",
  "ggplot2", "tidyverse", "ggh4x", "picante", "reshape2",
  "gridExtra", "grid", "lme4", "lmerTest", "broom.mixed"
)

missing <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(missing) > 0) {
  install.packages(missing, dependencies = TRUE)
}

# Load all libraries
invisible(lapply(required_packages, library, character.only = TRUE))
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

Load libraries and data
```{r}
library(phyloseq)
library(vegan)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(ggplot2)
library(tidyverse)
library(ggh4x)
library(picante)
library(reshape2)
library(gridExtra)
library(grid)

wd <- getwd()
file_path <- file.path(wd, "data", "molecular_ecology_17695.rds")
Cats <- readRDS(file_path)
Cats_ps <- Cats$phyloseq
sample_df <- sample_data(Cats_ps) %>% data.frame()
```


Table 1: Sample counts per location and origin

```{r}
table1 <- sample_df %>% 
  count(location, origin) %>%
  pivot_wider(names_from = origin, values_from = n, values_fill = 0)
knitr::kable(table1, caption = "Table 1: Number of samples by location and origin")

```

Figure 2: Relative abundance bar plot (genus level)

```{r}
ps_rel <- transform_sample_counts(Cats_ps, function(x) x / sum(x))
ps_genus <- tax_glom(ps_rel, taxrank = "genus")
plot_bar(ps_genus, fill = "genus") +
  facet_nested(~ location + origin, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_blank(), legend.position = "none") +
  labs(title = "Figure 2: Relative abundance of genera", 
       x = "Sample (by location and origin)", y = "Relative abundance")

```

Figure 3: MAG completness vs comtamination

```{r}
Cats_mag <- Cats$mag_info
ggplot(Cats_mag, aes(x = contamination, y = completeness)) +
  geom_point() +
  xlim(0, 8) + ylim(0, 100) +
  labs(x = "Contamination (%)", y = "Completeness (%)",
       title = "Figure 3: Scatterplot of MAG quality") +
  theme_bw()

```

Alpha diversity

Calculate Shannon, Observed richnedd and Faith's PD

```{r}
ps_pa <- transform_sample_counts(Cats_ps, function(x) ifelse(x > 0, 1, 0))

shannon <- estimate_richness(Cats_ps, measures = "Shannon") %>%
  rownames_to_column("sample_id")
observed <- estimate_richness(ps_pa, measures = "Observed") %>%
  rownames_to_column("sample_id")

otu_mat <- as(otu_table(Cats_ps), "matrix")
if (taxa_are_rows(Cats_ps)) otu_mat <- t(otu_mat)
pd_val <- pd(otu_mat, phy_tree(Cats_ps), include.root = FALSE) %>%
  rownames_to_column("sample_id") %>% select(sample_id, PD, SR)

alpha_all <- inner_join(shannon, observed, by = "sample_id") %>%
  inner_join(pd_val, by = "sample_id") %>%
  inner_join(sample_df %>% rownames_to_column("sample_id"), by = "sample_id")

alpha_long <- alpha_all %>%
  select(sample_id, location, origin, Shannon, Observed, PD) %>%
  pivot_longer(cols = c(Shannon, Observed, PD), names_to = "metric", values_to = "value")

```

Table 2: Kruskal-Wallis p-valuse by location and origin
```{r}
kruskal_loc <- alpha_long %>%
  group_by(metric) %>%
  summarise(p_location = kruskal.test(value ~ location)$p.value)
kruskal_orig <- alpha_long %>%
  group_by(metric) %>%
  summarise(p_origin = kruskal.test(value ~ origin)$p.value)
table2 <- inner_join(kruskal_loc, kruskal_orig, by = "metric")
knitr::kable(table2, caption = "Table 2: Kruskal‑Wallis p‑values for alpha diversity metrics", digits = 4)
```

Figure 4: Boxplot comparison across methods
```{r}
ggplot(alpha_long, aes(x = location, y = value)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  facet_wrap(~ metric, scales = "free_y") +
  theme_bw() +
  labs(y = "Alpha diversity", x = "Location", title = "Figure 4: Alpha diversity by location")
``` 


Beta diversity

Distance matrices
```{r}
bray_dist <- distance(ps_rel, method = "bray")
jaccard_dist <- distance(ps_pa, method = "jaccard")
unifrac_dist <- distance(ps_pa, method = "unifrac", weighted = FALSE)
wunifrac_dist <- distance(ps_rel, method = "unifrac", weighted = TRUE)
```

PCoA ordinations and variance explained (Table 3)
```{r}
ord_bray <- ordinate(ps_rel, method = "PCoA", distance = bray_dist)
ord_jaccard <- ordinate(ps_pa, method = "PCoA", distance = jaccard_dist)
ord_unifrac <- ordinate(ps_pa, method = "PCoA", distance = unifrac_dist)
ord_wunifrac <- ordinate(ps_rel, method = "PCoA", distance = wunifrac_dist)

get_var_exp <- function(ord) {
  eig <- ord$values$Eigenvalues
  var_exp <- eig / sum(eig)
  data.frame(Axis = paste0("Axis.", 1:2), 
             Variance = round(var_exp[1:2], 4),
             Cumulative = round(cumsum(var_exp)[1:2], 4))
}
var_df <- rbind(
  cbind(Method = "Bray-Curtis", get_var_exp(ord_bray)),
  cbind(Method = "Jaccard", get_var_exp(ord_jaccard)),
  cbind(Method = "Unweighted UniFrac", get_var_exp(ord_unifrac)),
  cbind(Method = "Weighted UniFrac", get_var_exp(ord_wunifrac))
)
knitr::kable(var_df, caption = "Table 3: Variance explained by first two PCoA axes")
```

Figure 5: Dispersion plots by origin (2x2 layout)
``` {r}
disp_bray_origin <- betadisper(bray_dist, sample_df$origin)
disp_jaccard_origin <- betadisper(jaccard_dist, sample_df$origin)
disp_unifrac_origin <- betadisper(unifrac_dist, sample_df$origin)
disp_wunifrac_origin <- betadisper(wunifrac_dist, sample_df$origin)

par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(disp_bray_origin, main = "Bray-Curtis: origin")
plot(disp_jaccard_origin, main = "Jaccard: origin")
plot(disp_unifrac_origin, main = "Unweighted UniFrac: origin")
plot(disp_wunifrac_origin, main = "Weighted UniFrac: origin")
par(mfrow = c(1, 1))
```

Figure 6: Dispersion plots by location (2x2 layout)
``` {r}
disp_bray_loc <- betadisper(bray_dist, sample_df$location)
disp_jaccard_loc <- betadisper(jaccard_dist, sample_df$location)
disp_unifrac_loc <- betadisper(unifrac_dist, sample_df$location)
disp_wunifrac_loc <- betadisper(wunifrac_dist, sample_df$location)

par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(disp_bray_loc, main = "Bray-Curtis: location")
plot(disp_jaccard_loc, main = "Jaccard: location")
plot(disp_unifrac_loc, main = "Unweighted UniFrac: location")
plot(disp_wunifrac_loc, main = "Weighted UniFrac: location")
par(mfrow = c(1, 1))
```
Figure 7: NMDS ordinations (2x2)
``` {r}
ord_nmds_bray <- ordinate(ps_rel, method = "NMDS", distance = bray_dist)
ord_nmds_jaccard <- ordinate(ps_pa, method = "NMDS", distance = jaccard_dist)
ord_nmds_unifrac <- ordinate(ps_pa, method = "NMDS", distance = unifrac_dist)
ord_nmds_wunifrac <- ordinate(ps_rel, method = "NMDS", distance = wunifrac_dist)

p1 <- plot_ordination(ps_rel, ord_nmds_bray, color = "location", shape = "origin") +
  geom_point(size = 3) + theme_bw() + ggtitle("Bray-Curtis")
p2 <- plot_ordination(ps_pa, ord_nmds_jaccard, color = "location", shape = "origin") +
  geom_point(size = 3) + theme_bw() + ggtitle("Jaccard")
p3 <- plot_ordination(ps_pa, ord_nmds_unifrac, color = "location", shape = "origin") +
  geom_point(size = 3) + theme_bw() + ggtitle("Unweighted UniFrac")
p4 <- plot_ordination(ps_rel, ord_nmds_wunifrac, color = "location", shape = "origin") +
  geom_point(size = 3) + theme_bw() + ggtitle("Weighted UniFrac")

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

PERMANOVA results (reported in text)
``` {r}
adonis_bray_loc <- adonis2(bray_dist ~ location, data = sample_df, permutations = 999)
adonis_bray_orig <- adonis2(bray_dist ~ origin, data = sample_df, permutations = 999)
adonis_jaccard_loc <- adonis2(jaccard_dist ~ location, data = sample_df, permutations = 999)
adonis_unifrac_loc <- adonis2(unifrac_dist ~ location, data = sample_df, permutations = 999)
adonis_wunifrac_loc <- adonis2(wunifrac_dist ~ location, data = sample_df, permutations = 999)

cat("Bray-Curtis ~ location:\n"); print(adonis_bray_loc)
cat("\nBray-Curtis ~ origin:\n"); print(adonis_bray_orig)
cat("\nJaccard ~ location:\n"); print(adonis_jaccard_loc)
cat("\nUnweighted UniFrac ~ location:\n"); print(adonis_unifrac_loc)
cat("\nWeighted UniFrac ~ location:\n"); print(adonis_wunifrac_loc)
```

Figure 8: Pairwise PERMANOVA matrices (2x2)
```{r}
pairwise_permanova_matrix <- function(dist_obj, metadata, group_var, groups_keep) {
  metadata <- metadata[rownames(metadata) %in% rownames(as.matrix(dist_obj)), ]
  meta_sub <- metadata %>% filter(.data[[group_var]] %in% groups_keep) %>% droplevels()
  dist_mat <- as.matrix(dist_obj)
  dist_sub <- dist_mat[rownames(meta_sub), rownames(meta_sub)]
  groups <- unique(meta_sub[[group_var]])
  results <- data.frame()
  for (i in 1:(length(groups)-1)) {
    for (j in (i+1):length(groups)) {
      g1 <- groups[i]; g2 <- groups[j]
      pair_meta <- meta_sub %>% filter(.data[[group_var]] %in% c(g1, g2))
      pair_samples <- rownames(pair_meta)
      pair_dist <- as.dist(dist_sub[pair_samples, pair_samples])
      perm <- adonis2(as.formula(paste("pair_dist ~", group_var)), data = pair_meta, permutations = 999)
      results <- rbind(results, data.frame(group1 = g1, group2 = g2, R2 = perm$R2[1], p = perm$`Pr(>F)`[1]))
    }
  }
  results$p_adj <- p.adjust(results$p, method = "BH")
  mat <- matrix(NA, nrow = length(groups), ncol = length(groups), dimnames = list(groups, groups))
  for (k in 1:nrow(results)) {
    mat[results$group1[k], results$group2[k]] <- results$p_adj[k]
    mat[results$group2[k], results$group1[k]] <- results$p_adj[k]
  }
  return(mat)
}

locations <- c("Spain", "Brazil", "Malaysia", "Denmark", "CaboVerde", "Aruba")
mat_bray <- pairwise_permanova_matrix(bray_dist, sample_df, "location", locations)
mat_jaccard <- pairwise_permanova_matrix(jaccard_dist, sample_df, "location", locations)
mat_unifrac <- pairwise_permanova_matrix(unifrac_dist, sample_df, "location", locations)
mat_wunifrac <- pairwise_permanova_matrix(wunifrac_dist, sample_df, "location", locations)

plot_matrix <- function(mat, title) {
  df <- melt(mat, na.rm = TRUE)
  ggplot(df, aes(Var1, Var2, fill = value)) +
    geom_tile() + geom_text(aes(label = round(value, 3))) +
    scale_fill_gradient(low = "white", high = "red") +
    theme_minimal() + labs(title = title, x = "", y = "", fill = "Adj p")
}
p_bray <- plot_matrix(mat_bray, "Bray-Curtis")
p_jaccard <- plot_matrix(mat_jaccard, "Jaccard")
p_unifrac <- plot_matrix(mat_unifrac, "Unweighted UniFrac")
p_wunifrac <- plot_matrix(mat_wunifrac, "Weighted UniFrac")

grid.arrange(p_bray, p_jaccard, p_unifrac, p_wunifrac, ncol = 2)
```
Preprocessing: relative abundance and long format

```{r}
## Preprocessing: relative abundance and long format
ps_rel <- transform_sample_counts(Cats_ps, function(x) x / sum(x))
otu_rel <- as(otu_table(ps_rel), "matrix")
if (taxa_are_rows(ps_rel)) otu_rel <- t(otu_rel)
otu_rel <- as.data.frame(otu_rel)

meta_rel <- as(sample_data(ps_rel), "data.frame")


tax_rel <- as.data.frame(tax_table(ps_rel))

# Long format for ggplot and per-taxon tests
dat_rel <- otu_rel %>%
  rownames_to_column("sample_id") %>%
  pivot_longer(-sample_id, names_to = "taxon", values_to = "rel_abund") %>%
  left_join(meta_rel %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  left_join(tax_rel %>% rownames_to_column("taxon"), by = "taxon")
```




Figure 9: Zero-inflation plot

```{r}
dat_rel %>%
  group_by(taxon) %>%
  summarise(prop_zero = mean(rel_abund == 0), .groups = "drop") %>%
  ggplot(aes(x = prop_zero)) +
  geom_histogram(binwidth = 0.05, fill = "steelblue", color = "black") +
  labs(x = "Proportion of samples with zero abundance", 
       y = "Number of taxa",
       title = "Figure 9: Zero inflation across taxa") +
  theme_bw()
```

Wilcoxon rank-sum test (domestic vs feral) per country

``` {r}
# Function to run Wilcoxon per country
run_wilcox_per_country <- function(country_name) {
  dat_rel %>%
    filter(location == country_name) %>%
    group_by(taxon, genus, species) %>%
    summarise(
      p_value = wilcox.test(rel_abund ~ origin, exact = FALSE)$p.value,
      mean_domestic = mean(rel_abund[origin == "Domestic"], na.rm = TRUE),
      mean_feral = mean(rel_abund[origin == "Feral"], na.rm = TRUE),
      log2_fc = log2((mean_domestic + 1e-6) / (mean_feral + 1e-6)),
      .groups = "drop"
    ) %>%
    mutate(q_value = p.adjust(p_value, method = "BH"),
           location = country_name)
}

countries <- unique(dat_rel$location)
wilcox_results_all <- map_dfr(countries, run_wilcox_per_country)

```
Figure 10: Top feral-enriched taxa per country

```{r}
wilcox_feral_enriched_all <- wilcox_results_all %>%
  group_by(location) %>%
  arrange(log2_fc, .by_group = TRUE) %>%
  slice_head(n = 5) %>%
  ungroup()

feral_plot_data <- wilcox_feral_enriched_all %>%
  select(location, taxon) %>%
  inner_join(dat_rel, by = c("location", "taxon")) %>%
  mutate(facet_label = paste0(location, ": ", taxon))

ggplot(feral_plot_data, aes(x = origin, y = rel_abund, fill = origin)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5) +
  facet_wrap(~ facet_label, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = c("Domestic" = "#E74C3C", "Feral" = "#3498DB")) +
  theme_bw() +
  labs(x = "Origin", y = "Relative abundance",
       title = "Figure 10: Top feral‑enriched taxa per country") +
  theme(strip.text = element_text(size = 8),
        axis.text.x = element_text(angle = 45, hjust = 1))
```


Figure 11: Top enriched taxa (domestic vs feral) per country
This figure shows, for each country, the five taxa with the smallest p‑value from a Wilcoxon test comparing domestic vs feral cats.
```{r}
wilcox_domestic_enriched_all <- wilcox_results_all %>%
  group_by(location) %>%
  arrange(desc(log2_fc), .by_group = TRUE) %>%
  slice_head(n = 5) %>%
  ungroup()

domestic_plot_data <- wilcox_domestic_enriched_all %>%
  select(location, taxon) %>%
  inner_join(dat_rel, by = c("location", "taxon")) %>%
  mutate(facet_label = paste0(location, ": ", taxon))

ggplot(domestic_plot_data, aes(x = origin, y = rel_abund, fill = origin)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5) +
  facet_wrap(~ facet_label, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = c("Domestic" = "#E74C3C", "Feral" = "#3498DB")) +
  theme_bw() +
  labs(x = "Origin", y = "Relative abundance",
       title = "Figure 11: Top domestic‑enriched taxa per country") +
  theme(strip.text = element_text(size = 8),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

Figure 12: Vocano plot: log2 fold change vs significance (per country)
This volcano plot displays, for each country, every taxon tested (domestic vs feral).

x‑axis: log₂ fold change (positive = higher in domestic cats, negative = higher in feral cats).

y‑axis: –log₁₀(p‑value) from Wilcoxon test.

Dashed horizontal line: significance threshold (p = 0.05).

Points in red are significant (p < 0.05).
The facet layout helps identify countries with stronger domestication signals (e.g., Brazil in the original report).
Note: The code adds a small pseudocount (1e-6) to avoid division by zero when calculating fold changes.

```{r}
wilcox_results_all %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & log2_fc > 0 ~ "Domestic-enriched",
      q_value < 0.05 & log2_fc < 0 ~ "Feral-enriched",
      TRUE ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = log2_fc, y = neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 1.5) +
    scale_color_manual(values = c("Domestic-enriched" = "#E74C3C", 
                                  "Feral-enriched" = "#3498DB", 
                                  "Not significant" = "grey60")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    facet_wrap(~ location, scales = "free") +
    labs(x = "Log2 fold change (Domestic vs Feral)", 
         y = "-log10(q-value)", 
         color = NULL,
         title = "Figure 12: Volcano plot – domestic vs feral cats per country") +
    theme_bw() +
    theme(strip.background = element_rect(fill = "lightgrey"))
```

CLR tranformation and RDA

```{r}
# CLR transformation (pseudocount to avoid log(0))
cat_data_clr <- transform_sample_counts(Cats_ps, function(x) {
  x <- x + 1e-9
  log(x) - mean(log(x))
})

otu_clr <- as(otu_table(cat_data_clr), "matrix")
if (taxa_are_rows(cat_data_clr)) otu_clr <- t(otu_clr)
otu_clr <- as.data.frame(otu_clr)
meta_clr <- as(sample_data(cat_data_clr), "data.frame")

# RDA: effect of origin while controlling for location
rda_model <- rda(otu_clr ~ origin + Condition(location), data = meta_clr)
```


Figure 13: RDA1 loadnings (top taxa driving domestic/feral separation)
```{r}
# Extract RDA1 loadings for taxa
taxon_scores <- scores(rda_model, display = "species", choices = 1)
rda1_loading <- as.data.frame(taxon_scores) %>%
  rownames_to_column("taxon") %>%
  rename(RDA1 = 2) %>%
  mutate(abs_load = abs(RDA1)) %>%
  arrange(desc(abs_load))

# Add taxonomy
tax_df <- as.data.frame(tax_table(Cats_ps)) %>%
  rownames_to_column("taxon") %>%
  select(taxon, genus, species)

rda1_loading <- rda1_loading %>%
  left_join(tax_df, by = "taxon") %>%
  mutate(label = ifelse(is.na(species), taxon, paste(genus, species, sep = " ")),
         direction = ifelse(RDA1 > 0, "Domestic-associated", "Feral-associated"))

# Top 20
rda1_loading %>%
  head(20) %>%
  ggplot(aes(x = reorder(label, RDA1), y = RDA1, fill = direction)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("Domestic-associated" = "#E74C3C", 
                                 "Feral-associated" = "#3498DB")) +
    labs(x = NULL, y = "RDA1 loading", fill = NULL,
         title = "Figure 13: Top 20 taxa driving separation between domestic and feral cats") +
    theme_bw()
```

A country-focused volcano plot
```{r}
## Volcano plot: each country vs all others (CLR-transformed, adjusted for origin)

# Ensure CLR data exists
if (!exists("dat_clr")) {
  cat_data_clr <- transform_sample_counts(Cats_ps, function(x) {
    x <- x + 1e-9
    log(x) - mean(log(x))
  })
  otu_clr <- as(otu_table(cat_data_clr), "matrix")
  if (taxa_are_rows(cat_data_clr)) otu_clr <- t(otu_clr)
  otu_clr <- as.data.frame(otu_clr)
  meta_clr <- as(sample_data(cat_data_clr), "data.frame")
  
  # tax_rel should already exist from earlier preprocessing
  dat_clr <- otu_clr %>%
    rownames_to_column("sample_id") %>%
    pivot_longer(-sample_id, names_to = "taxon", values_to = "clr") %>%
    left_join(meta_clr %>% rownames_to_column("sample_id"), by = "sample_id") %>%
    left_join(tax_rel %>% rownames_to_column("taxon"), by = "taxon")
}

# Compare each location vs all others
locations <- unique(dat_clr$location)

clr_location_results <- map_dfr(locations, function(this_location) {
  dat_clr %>%
    mutate(
      location_group = ifelse(location == this_location, this_location, "Other"),
      location_group = factor(location_group, levels = c("Other", this_location)),
      origin = relevel(factor(origin), ref = "Domestic")
    ) %>%
    group_by(taxon, genus, species) %>%
    group_modify(~ {
      fit <- lm(clr ~ location_group + origin, data = .x)
      broom::tidy(fit)
    }) %>%
    ungroup() %>%
    filter(str_detect(term, "^location_group")) %>%   # FIXED: captures the target location coefficient
    mutate(target_location = this_location)
}) %>%
  group_by(target_location) %>%
  mutate(q_value = p.adjust(p.value, method = "BH")) %>%
  ungroup()

# Volcano plot
clr_location_results %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & estimate > 0 ~ "Enriched in target location",
      q_value < 0.05 & estimate < 0 ~ "Depleted in target location",
      TRUE ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = estimate, y = neg_log10_q, color = direction)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  facet_wrap(~ target_location, scales = "free_y") +
  scale_color_manual(values = c("Enriched in target location" = "#2c7bb6",
                                "Depleted in target location" = "#d7191c",
                                "Not significant" = "grey70")) +
  labs(x = "CLR estimate (target location vs all others)",
       y = "-log10(q-value)",
       color = NULL,
       title = "Volcano plot: each country compared to all others (CLR, adjusted for origin)") +
  theme_bw() +
  theme(strip.background = element_rect(fill = "lightgrey"))
```





