---
title: "Differential abundance analysis of microbiome data"
author: "Antton Alberdi"
email: "antton.alberdi@sund.ku.dk"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
[wiki]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions
[wiki-load]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-1-load-and-inspect-the-data
[wiki-composition]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-2-visualise-community-composition
[wiki-single-taxon]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-3-explore-individual-taxa
[wiki-diagnose]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-4-diagnose-the-data-before-testing
[wiki-wilcoxon]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-5-wilcoxon-rank-sum-test
[wiki-glm]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-6-age-difference-using-log-transformed-linear-models
[wiki-adjusted]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-7-adjust-the-linear-model-for-health-status
[wiki-fdr]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-8-control-false-discoveries-with-fdr
[wiki-clr]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-9-compositional-analysis-with-clr-transformation
[wiki-compare]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-10-compare-differential-abundance-methods
[wiki-pca]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-11-pca-of-clr-transformed-gut-profiles
[wiki-rda]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-12-constrained-ordination-with-rda
[wiki-read-results]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-13-how-to-read-the-practical-results

## Loading of required libraries

This Rmd is the executable version of the practical. Detailed explanations are in the [Week 21 practical wiki][wiki]; the notes here only state what each chunk does and what to inspect.

Package context: [1. Load and Inspect the Data][wiki-load].

```{r}
library(phyloseq)
library(vegan)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(ggplot2)
library(tidyverse)
library(ggh4x)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(microbiome)
```

## Load and inspect the data

Details: [1. Load and Inspect the Data][wiki-load]. Load the saved `phyloseq` object, derive the practical metadata variables, and confirm that the abundance table, taxonomy table and sample metadata are present.

```{r}
cat_data <- readRDS("/Users/rikkelissau/KU/KU - 2/MM/molecular_ecology_17695 (1).rds")
cat_data <- cat_data$phyloseq
cat_data


rank_names(cat_data) #Output: "division" "phylum"   "class"    "order"    "family"   "genus"    "species" 
sample_variables(cat_data) #Output: "origin"   "location"

#Fra Git:
sample_data(cat_data) #Output: Dataframe med Sites og subjects, origin og location værdier
tax_table(cat_data) #Output: Taxonomi med kolonner: Division, phylum, class, order og family
otu_table(cat_data) #Output: Operational taxonomic units (OTU), clusters of highly similar nucleotide sequences that likely represent one or a few closely-related organisms
```
Check that the expected metadata variables are present, especially `body_site`, `age` and `status`.

## Visualise community composition

Details: [2. Visualise Community Composition][wiki-composition]. Convert counts to relative abundance and plot genus-level composition across body site, age and health status.

```{r}
cat_data_rel <- transform_sample_counts(cat_data, function(x) x / sum(x))
otu_rel <- as(otu_table(cat_data_rel), "matrix") %>% t() %>% as.data.frame()
meta_rel <- as(sample_data(cat_data_rel), "matrix") %>% as.data.frame()
tax_rel <- as(tax_table(cat_data_rel), "matrix") %>% as.data.frame()

cat_data_genus <- tax_glom(cat_data_rel, taxrank = "genus")

plot_bar(cat_data_genus, fill = "genus") +
  facet_nested(~ location + origin, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_blank())
#Giver plots over hvilke bakterier vi arbejder med, samt "sammenligner" de forskellige sites og feral/domestic fra de forskelligt sites
```
Use the stacked bar plot to decide whether gut samples look distinct enough to analyse separately.

The hidden chunk below creates `dat_rel`, the long relative-abundance table used by the remaining relative-abundance analyses.

```{r, include=FALSE}
dat_rel <- otu_rel %>%
  rownames_to_column("sample_id") %>%
  pivot_longer(-sample_id, names_to = "taxon", values_to = "rel_abund") %>%
  left_join(meta_rel %>% as("data.frame") %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  left_join(tax_rel %>% rownames_to_column("taxon"), by = "taxon")


#Indsat for at finde de taxons som har mest variation og dermed er mest forskellige?
variation_table <- dat_rel %>%
  group_by(taxon) %>%
  summarise(var = sd(rel_abund), .groups = "drop") %>%
  arrange(desc(var))

```

# Univariate statistics

These sections analyse taxa one at a time. Extended explanations are in the linked wiki sections.

## Explore individual taxa

Details: [3. Explore Individual Taxa][wiki-single-taxon]. Change `one_taxon` to inspect another taxon before running formal tests.

```{r}
one_taxon <- unique(dat_rel$taxon)[18] #Indsat frem for "one_taxon <- dat_rel$taxon[]"
#37 er okat spændende
#103 er okat spændende

# Comparison between country sites
dat_rel %>%
  filter(taxon == one_taxon) %>%
  ggplot(aes(x = location, y = rel_abund, fill = location)) + #ændret aes() så der ikke er body site etc., men vores variable
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15) +
    scale_y_log10() +
    theme_bw()
#Dette plot forklare den relative abundance for de forskellige site locations. Dette betyder, at vi har alle vores prøver for vores sites.



```
Look for zeros, skew and raw age separation in the selected taxon.

## Diagnose the data before testing

Details: [4. Diagnose the Data Before Testing][wiki-diagnose]. The next three chunks diagnose zero-inflation, skewness and non-normality among gut taxa.

```{r}

dat_rel %>%
  group_by(taxon) %>% #Plottet laves ud fra det taxon man valgte i R markdown lige ovenfor "one_taxon <- dat_rel$taxon[3100]"
  summarise(prop_zero = mean(rel_abund == 0), .groups = "drop") %>%
  ggplot(aes(x = prop_zero)) +
  geom_histogram(binwidth = 0.05, fill = "steelblue", color = "white") +
  theme_bw()


#Ændret fra:
#dat_rel %>%
  #filter(location == "gut") %>%
  #group_by(taxon) %>%
  #summarise(prop_zero = mean(rel_abund == 0), .groups = "drop") %>%
  #ggplot(aes(x = prop_zero)) +
    #geom_histogram(binwidth = 0.05, fill = "steelblue", color = "white") +
    #labs(
      #x = "Proportion of samples where relative abundance = 0",
      #y = "Number of taxa",
      #title = "Zero-inflation: most taxa are absent from most samples"
    #) +
    #theme_bw()

```
High zero proportions mean many taxa have limited information for per-taxon testing.

```{r}
set.seed(42)
example_taxa <- dat_rel %>%
  filter(location == "Aruba",) %>% #ændret fra body_site til location
  group_by(taxon, genus, species) %>%
  summarise(prop_nonzero = mean(rel_abund > 0), .groups = "drop") %>%
  filter(prop_nonzero > 0.5, prop_nonzero < 0.9) %>%
  slice_sample(n = 6) %>%
  pull(taxon)

dat_rel %>%
  filter(location == "Aruba", taxon %in% example_taxa, rel_abund > 0) %>% #ændret fra body_site til location
  mutate(label = species) %>%
  ggplot(aes(x = rel_abund)) +
    geom_histogram(bins = 20, fill = "steelblue", color = "white") +
    facet_wrap(~ label, scales = "free") +
    labs(x = "Relative abundance (non-zero values only)",
         y = "Count",
         title = "Right-skewed distributions in taxa") +
    theme_bw()

#Kopier denne kode så den tager for alle vores country location sites;
example_taxa <- dat_rel %>%
  filter(location == "CaboVerde",) %>% #ændret fra body_site til location
  group_by(taxon, genus, species) %>%
  summarise(prop_nonzero = mean(rel_abund > 0), .groups = "drop") %>%
  filter(prop_nonzero > 0.5, prop_nonzero < 0.9) %>%
  slice_sample(n = 6) %>%
  pull(taxon)

dat_rel %>%
  filter(location == "CaboVerde", taxon %in% example_taxa, rel_abund > 0) %>% #ændret fra body_site til location
  mutate(label = species) %>%
  ggplot(aes(x = rel_abund)) +
    geom_histogram(bins = 20, fill = "steelblue", color = "white") +
    facet_wrap(~ label, scales = "free") +
    labs(x = "Relative abundance (non-zero values only)",
         y = "Count",
         title = "Right-skewed distributions in taxa") +
    theme_bw()




```
Use these histograms to see whether non-zero abundances are dominated by a few high values.


```{r}
dat_rel %>%
  filter(location == "Aruba", taxon %in% example_taxa) %>%
  mutate(label = species) %>%
  ggplot(aes(sample = rel_abund)) +
    stat_qq(size = 0.8, alpha = 0.6) +
    stat_qq_line(color = "red", linetype = "dashed") +
    facet_wrap(~ label, scales = "free") +
    labs(x = "Theoretical quantiles (normal)",
         y = "Sample quantiles",
         title = "Q–Q plots: departures from normality in gut taxa") +
    theme_bw()



dat_rel %>%
  filter(taxon %in% example_taxa) %>%
  mutate(label = species) %>%
  ggplot(aes(sample = rel_abund)) +
    stat_qq(size = 0.8, alpha = 0.6) +
    stat_qq_line(color = "red", linetype = "dashed") +
    facet_wrap(~ label, scales = "free") +
    theme_bw()

```
Use the Q-Q plots as a visual check of normality on raw relative abundances.

## Wilcoxon rank-sum test

Details: [5. Wilcoxon Rank-Sum Test][wiki-wilcoxon]. This chunk runs one Wilcoxon test per gut taxon, compares infants with adults and adds a BH-adjusted `q_value`.

```{r}
#Ændret til:
#hvor man kan sammenligne locarions mod hinanden
#wilcox_results <- dat_rel %>%
  #filter(location %in% c("Aruba", "CaboVerde" )) %>%
  #group_by(taxon, genus, species) %>%
  #summarise(
    #p_value = wilcox.test(rel_abund ~ location, exact = FALSE)$p.value,
    #mean_Aruba = mean(rel_abund[location == "Aruba"]), #Her bruger vi "mean_Aruba" som vi genbruger i det næste markdown felt også, så hvis man ændrer denne til en anden lokation, ås skal man lige huske at ændre de andre variablenavne i de andre markdowns
    #mean_CaboVerde = mean(rel_abund[location == "CaboVerde"]), #Her bruger vi "mean_CaboVerde" som vi genbruger i det næste markdown felt også, så hvis man ændrer denne til en anden lokation, ås skal man lige huske at ændre de andre variablenavne i de andre markdowns
    #log2_fc = log2((mean_Aruba + 1e-6) / (mean_CaboVerde + 1e-6)),
    #.groups = "drop"
  #) %>%
  #mutate(q_value = p.adjust(p_value, method = "BH"))

#ALTERNATIVT::::::::::::
#Her laves der en wilcox test på baggrund af hver country location, men undersøger så den statistiske forskel på feral og domesic i det land:
wilcox_results_Aruba <- dat_rel %>%
  filter(location == "Aruba") %>%
  group_by(taxon, genus, species) %>%
  summarise(
    p_value = wilcox.test(rel_abund ~ origin, exact = FALSE)$p.value,
    mean_domestic = mean(rel_abund[origin == "Domestic"]),
    mean_feral = mean(rel_abund[origin == "Feral"]),
    log2_fc = log2((mean_domestic + 1e-6) / (mean_feral + 1e-6)),
    .groups = "drop"
  ) %>%
  mutate(q_value = p.adjust(p_value, method = "BH"))

# Brazil
wilcox_results_Brazil <- dat_rel %>%
  filter(location == "Brazil") %>%
  group_by(taxon, genus, species) %>%
  summarise(
    p_value = wilcox.test(rel_abund ~ origin, exact = FALSE)$p.value,
    mean_domestic = mean(rel_abund[origin == "Domestic"]),
    mean_feral = mean(rel_abund[origin == "Feral"]),
    log2_fc = log2((mean_domestic + 1e-6) / (mean_feral + 1e-6)),
    .groups = "drop"
  ) %>%
  mutate(q_value = p.adjust(p_value, method = "BH"))

#Spain
wilcox_results_Spain <- dat_rel %>%
  filter(location == "Spain") %>%
  group_by(taxon, genus, species) %>%
  summarise(
    p_value = wilcox.test(rel_abund ~ origin, exact = FALSE)$p.value,
    mean_domestic = mean(rel_abund[origin == "Domestic"]),
    mean_feral = mean(rel_abund[origin == "Feral"]),
    log2_fc = log2((mean_domestic + 1e-6) / (mean_feral + 1e-6)),
    .groups = "drop"
  ) %>%
  mutate(q_value = p.adjust(p_value, method = "BH"))

# Denmark
wilcox_results_Denmark <- dat_rel %>%
  filter(location == "Denmark") %>%
  group_by(taxon, genus, species) %>%
  summarise(
    p_value = wilcox.test(rel_abund ~ origin, exact = FALSE)$p.value,
    mean_domestic = mean(rel_abund[origin == "Domestic"]),
    mean_feral = mean(rel_abund[origin == "Feral"]),
    log2_fc = log2((mean_domestic + 1e-6) / (mean_feral + 1e-6)),
    .groups = "drop"
  ) %>%
  mutate(q_value = p.adjust(p_value, method = "BH"))

wilcox_results_CaboVerde <- dat_rel %>%
  filter(location == "CaboVerde") %>%
  group_by(taxon, genus, species) %>%
  summarise(
    p_value = wilcox.test(rel_abund ~ origin, exact = FALSE)$p.value,
    mean_domestic = mean(rel_abund[origin == "Domestic"]),
    mean_feral = mean(rel_abund[origin == "Feral"]),
    log2_fc = log2((mean_domestic + 1e-6) / (mean_feral + 1e-6)),
    .groups = "drop"
  ) %>%
  mutate(q_value = p.adjust(p_value, method = "BH"))

# Malaysia
wilcox_results_Malaysia <- dat_rel %>%
  filter(location == "Malaysia") %>%
  group_by(taxon, genus, species) %>%
  summarise(
    p_value = wilcox.test(rel_abund ~ origin, exact = FALSE)$p.value,
    mean_domestic = mean(rel_abund[origin == "Domestic"]),
    mean_feral = mean(rel_abund[origin == "Feral"]),
    log2_fc = log2((mean_domestic + 1e-6) / (mean_feral + 1e-6)),
    .groups = "drop"
  ) %>%
  mutate(q_value = p.adjust(p_value, method = "BH"))





```
Positive `log2_fc` means adult-enriched; negative means infant-enriched. Interpret `q_value` using [8. Control False Discoveries with FDR][wiki-fdr].

The next table lists the strongest infant-enriched taxa.

```{r}

#wilcox_location_enriched <- wilcox_results %>%
  #select(taxon, genus, species, mean_Aruba, mean_CaboVerde, log2_fc, p_value, q_value) %>% #Her bruger vi mean_Aruba og CaboVerde, men alt efter 
  #arrange(log2_fc) %>% 
  #head(10)

#wilcox_location_enriched




# Kombiner alle wilcox_result dataframes
wilcox_results_all <- bind_rows(
  wilcox_results_Aruba %>% mutate(location = "Aruba"),
  wilcox_results_Brazil %>% mutate(location = "Brazil"),
  wilcox_results_Spain %>% mutate(location = "Spain"),
  wilcox_results_Denmark %>% mutate(location = "Denmark"),
  wilcox_results_CaboVerde %>% mutate(location = "CaboVerde"),
  wilcox_results_Malaysia %>% mutate(location = "Malaysia")
)

# Nu kan du lave top 10 Feral-berigede for hvert land samlet
wilcox_feral_enriched_all <- wilcox_results_all %>%
  group_by(location) %>%
  arrange(log2_fc, .by_group = TRUE) %>%
  slice_head(n = 5) %>%
  ungroup()

# Se resultatet
wilcox_feral_enriched_all %>%
  select(location, taxon, genus, species, mean_domestic, mean_feral, log2_fc, q_value)

#Her kan man så se ud fra kolonnen "log2_fc" hvorvidr det taxon (kol 1) er mere abundant i feral (hvis der er en negativ log2_fc) eller i domestic (positiv log2_fc)
#Her kan vi fx. se at i Aruba findes der bin_caboverde.10 hos 0 af de fdomestic katte, men findes i feral. Selvom det hedder CaboVerde kan det godt fidnes i Aruba, men det taxon blev dog først fundet i Caboverde
#Det første plot inkluderer kun dem som er i feral, og det andet er dem som findes i domestic
wilcox_domestic_enriched_all <- wilcox_results_all %>%
  group_by(location) %>%
  arrange(desc(log2_fc), .by_group = TRUE) %>%   # sorter faldende (størst positiv først)
  slice_head(n = 5) %>%
  ungroup()

# Se resultatet
wilcox_domestic_enriched_all %>%
  select(location, taxon, genus, species, mean_domestic, mean_feral, log2_fc, q_value)
```
Inspect the top infant-enriched taxon in the raw data before trusting the table.

```{r}
# Forbered data: Hent de rå abundanser for alle de taxa, der er i toppen
feral_plot_data <- wilcox_feral_enriched_all %>%
  # Behold kun de kolonner vi skal bruge til at filtrere
  select(location, taxon) %>%   # fjern species – den findes måske ikke
  # Hent de tilsvarende rækker fra dat_rel
  inner_join(dat_rel, by = c("location", "taxon")) %>%
  # Lav en label til facetter (location + taxon)
  mutate(facet_label = paste0(location, ": ", taxon))

# Tjek om der er data
head(feral_plot_data)

# Plot
ggplot(feral_plot_data, aes(x = origin, y = rel_abund, fill = origin)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5) +
  facet_wrap(~ facet_label, scales = "free_y", ncol = 4) +
  scale_fill_manual(values = c("Domestic" = "#E74C3C", "Feral" = "#3498DB")) +
  theme_bw() +
  labs(x = "Origin", y = "Relative abundance",
       title = "Top Feral-enriched taxa per country") +
  theme(strip.text = element_text(size = 8),
        axis.text.x = element_text(angle = 45, hjust = 1))



domestic_plot_data <- wilcox_domestic_enriched_all %>%
  select(location, taxon) %>%
  inner_join(dat_rel, by = c("location", "taxon")) %>%
  mutate(facet_label = paste0(location, ": ", taxon))

ggplot(domestic_plot_data, aes(x = origin, y = rel_abund, fill = origin)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5) +
  facet_wrap(~ facet_label, scales = "free_y", ncol = 4) +
  scale_fill_manual(values = c("Domestic" = "#E74C3C", "Feral" = "#3498DB")) +
  theme_bw() +
  labs(x = "Origin", y = "Relative abundance",
       title = "Top Domestic-enriched taxa per country") +
  theme(strip.text = element_text(size = 8),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
Check whether the visual direction matches the negative `log2_fc`.

The next table lists the strongest adult-enriched taxa.

```{r}
# Vis top 10 mest Feral-berigede taxa (lavest log2_fc = højere i Feral) for hvert land
wilcox_feral_enriched <- wilcox_results_all %>%
  group_by(location) %>%
  arrange(log2_fc, .by_group = TRUE) %>%
  slice_head(n = 10) %>%
  ungroup()

# Se resultatet
wilcox_feral_enriched %>%
  select(location, taxon, genus, species, mean_domestic, mean_feral, log2_fc, p_value, q_value)
```
Inspect the top adult-enriched taxon in the raw data.

```{r}
# Mest Domestic-berigede (højeste log2_fc) per land
top_domestic_per_country <- wilcox_results_all %>%
  group_by(location) %>%
  arrange(desc(log2_fc)) %>%
  slice_head(n = 1) %>%
  ungroup()

# Mest Feral-berigede (laveste log2_fc) per land
top_feral_per_country <- wilcox_results_all %>%
  group_by(location) %>%
  arrange(log2_fc, .by_group = TRUE) %>%
  slice_head(n = 1) %>%
  ungroup()


# Forudsætning: top_domestic_per_country og top_feral_per_country findes
# (som tidligere defineret)

# Liste over lande
countries <- unique(wilcox_results_all$location)

# 1. Loop over Domestic-berigede taxa (én pr. land)
for (land in countries) {
  dom_taxon <- top_domestic_per_country %>% filter(location == land) %>% pull(taxon)
  
  p_dom <- dat_rel %>%
    filter(location == land, taxon == dom_taxon) %>%
    ggplot(aes(x = origin, y = rel_abund, fill = origin)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(width = 0.15, alpha = 0.6) +
      scale_fill_manual(values = c("Domestic" = "#E74C3C", "Feral" = "#3498DB")) +
      theme_bw() +
      labs(title = paste(land, "- Most Domestic-enriched taxon"),
           x = "Origin", y = "Relative abundance")
  
  print(p_dom)
  # Gem evt. med: ggsave(paste0("Domestic_", land, ".png"), p_dom, width = 6, height = 4)
}
```
Check whether the visual direction matches the positive `log2_fc`.

The volcano plot combines effect size and adjusted significance for the Wilcoxon results.

```{r}
# Volcano plot per country
wilcox_results_all %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & log2_fc > 0 ~ "Domestic-enriched",
      q_value < 0.05 & log2_fc < 0 ~ "Feral-enriched",
      TRUE ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = log2_fc, y = neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 1.5) +
    scale_color_manual(values = c("Domestic-enriched" = "#E74C3C", 
                                  "Feral-enriched" = "#3498DB", 
                                  "Not significant" = "grey60")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    facet_wrap(~ location, scales = "free") +
    labs(x = "Log2 fold change (Domestic vs Feral)", 
         y = "-log10(q-value)", 
         color = NULL,
         title = "Wilcoxon rank test: Domestic vs Feral effect per country") +
    theme_bw() +
    theme(strip.background = element_rect(fill = "lightgrey"))



```
Use the volcano plot to identify taxa that are both large-effect and statistically supported.

The next plot shows Wilcoxon effect size on the x-axis and orders taxa by q-value.

```{r}
# Top 5 Domestic- og Feral-berigede per land
top_per_land <- wilcox_results_all %>%
  group_by(location) %>%
  slice_max(log2_fc, n = 5, with_ties = FALSE) %>%
  bind_rows(
    wilcox_results_all %>%
      group_by(location) %>%
      slice_min(log2_fc, n = 5, with_ties = FALSE)
  ) %>%
  mutate(
    label = ifelse(is.na(species), taxon, species),
    direction = ifelse(log2_fc > 0, "Domestic-enriched", "Feral-enriched")
  )

ggplot(top_per_land, aes(x = reorder(label, log2_fc), y = log2_fc, fill = direction)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ location, scales = "free_y") +
  scale_fill_manual(values = c("Domestic-enriched" = "#E74C3C", "Feral-enriched" = "#3498DB")) +
  labs(x = NULL, y = "Log2 fold change (Domestic vs Feral)", fill = NULL,
       title = "Top differentially abundant taxa per country (Wilcoxon)") +
  theme_bw() +
  theme(strip.text = element_text(size = 10))
```
Use the bar chart as a first-pass rank list, then compare it with the model-based analyses below.



## Compositional analysis with CLR transformation

Details: [9. Compositional Analysis with CLR Transformation][wiki-clr]. Transform counts with CLR before fitting the adjusted age model on a compositional scale.

```{r}
## CLR transformation on cat_data (counts)
cat_data_clr <- transform_sample_counts(cat_data, function(x) {
  x <- x + 1e-9   # pseudo-count to avoid log(0)
  log(x) - mean(log(x))
})

# Extract CLR matrix: samples as rows, taxa as columns
otu_clr <- as(otu_table(cat_data_clr), "matrix")
if (taxa_are_rows(cat_data_clr)) otu_clr <- t(otu_clr)
otu_clr <- as.data.frame(otu_clr)

# Metadata
meta_clr <- as(sample_data(cat_data_clr), "data.frame")
```
The hidden chunk below reshapes CLR values into the long table `dat_clr`.


# Multivariate statistics

These sections analyse all gut taxa jointly on the CLR scale. PCA is unconstrained; RDA is constrained by metadata.

## PCA of CLR-transformed gut profiles

Details: [11. PCA of CLR-Transformed Gut Profiles][wiki-pca]. Build a gut-only CLR matrix and inspect the largest unconstrained community gradients.

```{r}
## PCA of CLR-transformed data (all samples, no filtering)
pca <- prcomp(otu_clr, center = TRUE, scale. = FALSE)
pca_var <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)

pca_scores <- as.data.frame(pca$x[, 1:2]) %>%
  rownames_to_column("sample_id") %>%
  left_join(meta_clr %>% rownames_to_column("sample_id"), by = "sample_id")

ggplot(pca_scores, aes(x = PC1, y = PC2, color = origin, shape = location)) +
  geom_point(size = 3, alpha = 0.85) +
  scale_color_manual(values = c("Domestic" = "#E74C3C", "Feral" = "#3498DB")) +
  labs(x = paste0("PC1 (", pca_var[1], "%)"),
       y = paste0("PC2 (", pca_var[2], "%)"),
       color = "Origin", shape = "Location",
       title = "PCA of CLR-transformed cat microbiome") +
  theme_bw()
```
Use PCA to see whether age or health status align with the strongest overall community gradients.

## Constrained ordination with RDA

Details: [12. Constrained Ordination with RDA][wiki-rda]. Fit an RDA on the gut CLR matrix constrained by age and health status.

```{r}
## RDA: effect of origin while controlling for location
rda_model <- rda(otu_clr ~ origin + Condition(location), data = meta_clr)

# Permutation tests
anova(rda_model, permutations = 999)          # whole model
anova(rda_model, by = "term", permutations = 999)  # test of origin alone

# (Optional) explained variance
rda_var <- round(100 * rda_model$CCA$eig[1] / rda_model$tot.chi, 1)
```


```{r}
# Extract RDA1 loadings for taxa
taxon_scores <- scores(rda_model, display = "species", choices = 1)
rda1_loading <- as.data.frame(taxon_scores) %>%
  rownames_to_column("taxon") %>%
  rename(RDA1 = 2) %>%
  mutate(abs_load = abs(RDA1)) %>%
  arrange(desc(abs_load))

# Add taxonomy
tax_df <- as.data.frame(tax_table(cat_data)) %>%
  rownames_to_column("taxon") %>%
  select(taxon, genus, species)

rda1_loading <- rda1_loading %>%
  left_join(tax_df, by = "taxon") %>%
  mutate(label = ifelse(is.na(species), taxon, paste(genus, species, sep = " ")),
         direction = ifelse(RDA1 > 0, "Domestic-associated", "Feral-associated"))

# Top 20
rda1_loading %>%
  head(20) %>%
  ggplot(aes(x = reorder(label, RDA1), y = RDA1, fill = direction)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("Domestic-associated" = "#E74C3C", "Feral-associated" = "#3498DB")) +
    labs(x = NULL, y = "RDA1 loading", fill = NULL,
         title = "Top 20 taxa driving separation between Domestic and Feral cats") +
    theme_bw()
```
